### 2015-05-11 whoop.tex
fix

We now formalize our approach to verifying data race freedom in a driver by statically analyzing its locksets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronization bugs that arise from careless fine-grained locking. The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.

@@

Another limitation of \whoop is that it is unable to verify cases where a driver is forced to be accessed by a single process at a time. This \emph{single-open device}~\cite{corbet2005linux} mode is enforced by using an atomic counter to indicate device availability: an entry point executes if the counter has value 1, else blocks. Because \whoop performs pair-wise analysis, it over-approximates this counter, and regard two entry points as running concurrently, even if this is not the case. However,~\cite{corbet2005linux} advises against using single-open device mode, as it hinders user ingenuity (e.g. a user might expect that a device can be accessed concurrently for performance). Thus, because using single-open device mode is considered bad practice, we burden the developer with disregarding any related false bugs.

>>>>

We now formalize our approach to verifying data race freedom in a driver by statically analyzing its locksets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronization bugs that arise from careless fine-grained locking. Note that the actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.

@@

Another limitation of \whoop is that it is unable to verify drivers that are designed to be accessed by a single process at a time. This \emph{single-open device}~\cite{corbet2005linux} mode can be enforced by atomically testing (at the beginning of each entry point) a flag that indicates device availability: if the flag is set to true, then the respective entry point executes, else it blocks. Because \whoop performs pair-wise analysis, it over-approximates this flag, and regards a pair of entry points as running concurrently, even if this is not the case. However,~\cite{corbet2005linux} advises against serializing drivers in this way, as it hinders user ingenuity (e.g. a user might expect that a device can be accessed concurrently for performance). Because using single-open device mode is considered bad practice, we thus burden the developer with disregarding any related false bug reports.
